---
title: "data_science_day1"
output: html_document
date: "2024-01-18"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1.4 Data Basics in RStudio

```{r}
# view(gm) allows you to view the data

# head(gm) allows you to view the top six rows of the data 

# glimpse(gm) gives a summary on the rows, columns, and types of variables in the dataset (as well as a "glimpse" of the types of entries in each variable)

# dim(gm) gives the dimensions of the dataset in vector format [rows cols]

# names(gm) gives the names of all the variables in the set
```


# 1.7 Codebooks
```{r}
# code to view codebook in rstudio, if applicable
?dslabs::gapminder
```
a) infant_mortality measures infant deaths per 1000
b) 1 per thousand 

# 1.8 Examining Single Variable 
a) using gm$fertility notation (where first is dataset, and second is the variable)
b) class(gm$feritlity) notation to understand format/class of variable 
c) levels(gm$continent) determines levels 

## 2 R Markdown and Reproducible Research

a) bullets using a star , italics using surrounding stars *hello*, section headers using # 
b) R code appears within backticks 
c) HTML webpage has the r code, as well as the outputs 


## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
